# Anatomy of a ggplot {#sec-anatomy}

You can get really far with ggplot2 by copying an example, changing the variable names, and a bit of use of an LLM.

One of the challenges with this, is that you might not have internalised the mental model of how ggplot plots are built. The system is rich, and built on a great foundation of theory that help you think about both how to build your data for data analysis, and how to then present it in a plot. 

This chapter is about the underlying features of ggplot2. We will take plots, tease them apart, and put them back together.

::: {.overview}

## Overview {.unnumbered}

**Duration** 45 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What is this plot telling me, and what is it made of?
- What are the seven pieces of a ggplot?
- How do I build one up from nothing?
- How do I map a variable onto colour, shape, or size?
- Why does `colour` inside `aes()` do something different to `colour` outside it?
- How do I add a second layer?

:::

```{r}
#| label: setup
#| message: false
library(ggplot2)
library(naniar)
```

## A nice looking plot

> Here's one I prepared earlier.

Let's start with a mostly finished product. Let's take a moment to just look at it.

```{r}
#| label: fig-elnino
#| warning: false
#| fig-alt: |
#|   Scatterplot of sea temperature against air temperature. The points form
#|   two separate clusters. The 1993 readings sit between 22 and 25 degrees,
#|   and the 1997 readings sit between 26 and 30 degrees.

ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    colour = factor(year)
  )
) +
  geom_point(alpha = 0.6)
```

Each point is a reading from a buoy floating in the tropical Pacific. The
horizontal axis is the air temperature, the vertical axis is the temperature of
the sea underneath it, and the colour is the year the reading was taken.

There are two clouds of data points.

1. One between 20-25C
2. One up between 26-29C

So what happened between 1993 and 1997?

```{r}
#| label: mean-temps
library(dplyr)
oceanbuoys |> 
  group_by(year) |> 
  summarise(
    mean_sea_temp_c = mean(sea_temp_c, na.rm = TRUE)
  )
```

It is a big jump - nearly five degrees celcius, in the same patch of ocean, four years apart. This is the result of
[El Niño](https://en.wikipedia.org/wiki/El_Ni%C3%B1o), and the 1997 to 1998
event was one of the strongest on record.

I think that is a lovely thing to be able to see in a plain scatterplot.

Now let's remove the colour, and see how that clear story disappears.

```{r}
#| label: fig-elnino-nocolour
#| warning: false
#| fig-alt: |
#|   The same scatterplot without colour. The points now read as one elongated
#|   cloud with a gap in the middle.
#| fig-cap: "The same readings, with the year removed."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
) +
  geom_point(alpha = 0.6)
```

You can still see this gap in the data, but you lose some of that strong clarity.

The gap is the story! And the plot showed this nicely. We mapped the variable "year" onto "colour", and that made this clearer.

A big question when showing data is to think:

> What is the right plot to use

Which is a good question.

But now I think we can also add to this:

> What do I want the plot to tell me? And what have I left out?

### What is it made of?

Let's take @fig-elnino apart.

Let's take another look at the code

1. We see **data**, `oceanbuoys`, which is from the {naniar} package.
2. There is a set of instructions saying which column goes where. `air_temp_c`
across, `sea_temp_c` up, `year` onto colour. Those live inside `aes()`.
3. And there is a `+` between them, which is how you glue the parts together.
4. There is something saying **draw a point for each row**: is `geom_point()`.

```{r}
library(naniar)
library(ggplot2)
ggplot(
  data = oceanbuoys,      # <1>
  aes(                    # <2>
    x = air_temp_c,       # <2>
    y = sea_temp_c,       # <2>
    color = factor(year)  # <2>
  )
) +                       # <3>
  geom_point(alpha = 0.6) # <4>
```


::: {.callout-note title="Your Turn"}

Look at @fig-elnino and answer these without running any code.

1. How many rows do you think `oceanbuoys` has? Have a guess, then check with
   `nrow(oceanbuoys)`.
2. If you mapped `humidity` onto the y axis instead of `sea_temp_c`,
   which part of the code would change?
3. If you wanted crosses instead of dots, which part would change?

:::

## The seven pieces

The grammar of graphics is an idea from
[Leland Wilkinson](https://en.wikipedia.org/wiki/Leland_Wilkinson), and ggplot2
is [Hadley Wickham's](https://hadley.nz/) implementation of it. It is described as a grammar, in the sense that you can express complex things with few words to say something complex.

There are seven parts to the grammar of graphics - and you don't need to know them all by heart. But they are very useful pieces of the map to have.

| Piece | Its job | Chapter |
|---|---|---|
| **Data** | The data frame you are plotting | @sec-shape |
| **Aesthetics** | Which column maps onto which visual channel | this chapter |
| **Geoms** | What mark gets drawn | this chapter |
| **Stats** | What gets computed before drawing | @sec-computed |
| **Scales** | How data values turn into positions and colours | @sec-computed |
| **Coords** | The coordinate system the marks sit in | @sec-computed |
| **Facets** | How the plot splits into panels | @sec-comparing |

In this chapter we focus on using three of them: Data, aesthetics, and geoms.

We will come back to this table as we go in the course. But I find that knowing the names of these features of the plot provide useful features of a mental model to focus on and understand.

## Building a plot back up

We took a plot apart before. Let's build one up again.

Start with nothing but the data.

```{r}
#| label: fig-blank
#| fig-alt: A completely blank grey panel with no axes and no marks.
#| fig-cap: "Data, and nothing else."
ggplot(data = oceanbuoys)
```

This isn't super inspiring, but it is worth thinking about. I told ggplot2
what data to use, and absolutely nothing else. It does not know what to put on
the axes, and it does not know what to draw. It gives you a panel and waits.

Now let's say which columns go where, by specifying how they map onto an axes:

```{r}
#| label: fig-axes
#| fig-alt: |
#|   A grey panel with labelled axes for air temperature and sea temperature,
#|   but no points.
#| fig-cap: "Now with aesthetics. Axes appear, but nothing is drawn."
ggplot(
  data = oceanbuoys,
  mapping = aes(
    x = air_temp_c,
    y = sea_temp_c
  )
)
```

Now we get axes! These correspond to the range of the data. But we haven't said what data to put on there. We specify that with **A geom**.

```{r}
#| label: fig-points
#| warning: false
#| fig-alt: |
#|   The scatterplot with black points, forming one elongated cloud with a gap
#|   in the middle.
#| fig-cap: "And now a geom. This is a plot."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
) +
  geom_point()
```

A plot!

Three lines, three pieces. Data, aesthetics, geom.

Let's show this as a set of panel tabs, to get a bit more of a sense of the plot being built up.


:::{.panel-tabset}

## grey

```{r}
ggplot(oceanbuoys)
```

## axes

```{r}
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
)
```

## geom

```{r}
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
) +
  geom_point()
```


:::

### A small (but worthwhile) digression into missing data

You might have noticed a warning:

```
Removed 81 rows containing missing values or values outside the scale range
```

That is no mistake! There are 81 readings in this data where the air
temperature was never recorded, and ggplot2 cannot draw a point when it does not
know where to put it, so it drops them and tells you.

This is a good thing, I think. For two reasons:

1. The fact it told you at all! Missing values are often omitted. It's nice ggplot2 is honest.

2. The plot you are looking at is __missing__ 81 readings, and nothing on the plot itself shows you that. You would have to read the warning.

I've spent a lot of time looking at missing data, and so I wrote two R packages to help explore data, and missing data: [{visdat}](https://docs.ropensci.org/visdat/), and [{naniar}](https://naniar.njtierney.com/).

A plot that I think is worthwhile showing is the `vis_miss()` plot from `{visdat}`:

```{r}
library(visdat)
vis_miss(oceanbuoys)
```

This shows the overall missingness in the dataset.

We come back to missing data properly in @sec-shape, where we draw those rows instead of throwing them away. For now, just know that they are missing.

<!-- NOTE: UP TO HERE -->

### Colour

```{r}
#| label: fig-rebuilt
#| warning: false
#| fig-alt: |
#|   The scatterplot with points coloured by year, showing two separate
#|   clusters for 1993 and 1997.
#| fig-cap: "Mapping year onto colour. The gap becomes a story."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    colour = factor(year)
  )
) +
  geom_point(alpha = 0.6)
```

One extra aesthetic, and the plot went from a shrug to a finding.

::: {.callout-note title="Your Turn"}

Build a plot up the same way, one piece at a time. Run each line before you add
the next, so you can see what each part contributes.

1. `ggplot(oceanbuoys)`
2. Add `aes(x = humidity, y = air_temp_c)`
3. Add `geom_point()`
4. Now add `colour = factor(year)` to the `aes()`

What does the year do to this plot? Is it as dramatic as it was for sea
temperature?

:::

## Mapping more variables

Position is not the only channel available. There are several, and you map a
variable onto one the same way every time, inside `aes()`.

Let's try shape.

```{r}
#| label: fig-shape
#| warning: false
#| fig-alt: |
#|   Scatterplot where 1993 readings are circles and 1997 readings are
#|   triangles.
#| fig-cap: "Year mapped onto shape instead of colour."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    shape = factor(year)
  )
) +
  geom_point(alpha = 0.6)
```

It works, and I think it is clearly worse than the colour version.

The two groups are separated well enough here that you can still see them, but
you are now doing shape recognition instead of colour recognition, and shape is
slower. Try the same thing with five groups and it falls apart completely.

Now let's try size.

```{r}
#| label: fig-size
#| warning: false
#| fig-alt: |
#|   Scatterplot where point size varies with humidity, producing a cluttered
#|   overlapping mess.
#| fig-cap: "Humidity mapped onto size. This is not a good plot."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    size = humidity
  )
) +
  geom_point(alpha = 0.4)
```

Hmm.

There is information in there, technically. But the big points are covering the
small ones, the sizes are hard to compare unless they happen to be next to each
other, and the whole thing is harder to read than the plot we started with.

Size is nearly always worse than people expect. I reckon that is because it
looks like it comes for free, when in fact it costs you the thing that made the
plot readable in the first place.

**The channels are not equal.** Position is the best one by a long way, which is
why x and y do the heavy lifting in almost every good plot. Colour is a strong
second for a small number of groups. Shape works for a few categories. Size is
a last resort.

We come back to this properly in @sec-eye, where choosing between channels
becomes the whole question.

::: {.callout-note title="Your Turn"}

1. Make a scatterplot of `wind_ew` against `wind_ns`, coloured by year.
2. Now map `sea_temp_c` onto colour instead. What is different about the legend?
3. Try mapping `year` onto colour **without** wrapping it in `factor()`. What
   happens, and why do you think that is?

:::

## Inside and outside `aes()`

This one catches everybody, and it caught me for an embarrassingly long time.

Say you want blue points. Reasonable enough. So you write this.

```{r}
#| label: fig-bluebug
#| warning: false
#| fig-alt: |
#|   Scatterplot with salmon-coloured points and a legend titled "colour" with
#|   a single entry labelled "blue".
#| fig-cap: "Asking for blue. Getting salmon, and a legend."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    colour = "blue"
  )
) +
  geom_point()
```

The points are salmon, and there is a legend with a single entry, labelled
"blue".

So what happened?

Everything inside `aes()` is a **mapping**. You are saying "take this variable
and use it to decide the colour", so ggplot2 did exactly that. It took the value
`"blue"`, treated it as a variable with one single category, and gave that
category the first colour from its default palette, which happens to be salmon.

The legend is the real giveaway here. ggplot2 built one because it thought you
were mapping something, and legends are what you get when you map.

If you want to **set** a colour rather than map one, it goes outside `aes()`,
in the geom.

```{r}
#| label: fig-bluefixed
#| warning: false
#| fig-alt: Scatterplot with blue points and no legend.
#| fig-cap: "Setting the colour instead of mapping it."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
) +
  geom_point(colour = "blue")
```

Blue points, and no legend at all, because nothing was mapped.

::: {.callout-important title="This will bite you"}

The rule is short, and it is worth memorising.

**Inside `aes()` maps a variable. Outside `aes()` sets a value.**

If you get an unexpected legend, you have accidentally mapped something. If you
get an unexpected colour, look at which side of the bracket it is sitting on.

:::

::: {.callout-note title="Your Turn"}

1. Make the points in your wind plot dark green and slightly transparent.
2. Now deliberately put `colour = "darkgreen"` inside `aes()` and look at what
   you get.
3. `size = 3` and `alpha = 0.5` follow exactly the same rule. Try setting each
   of them in both places.

:::

## Layers

The `+` is doing real work, and it's worth being explicit about what.

Each thing you add with `+` is a **layer**, and layers draw on top of each other
in the order you write them. So let's add a second one.

```{r}
#| label: fig-smooth
#| message: false
#| warning: false
#| fig-alt: |
#|   Scatterplot coloured by year with a separate fitted trend line running
#|   through each of the two clusters.
#| fig-cap: "Two layers. Points, then a smooth on top."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    colour = factor(year)
  )
) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm")
```

Two trend lines, one per year.

I never told `geom_smooth()` about the years. It inherited the whole `aes()`
from the `ggplot()` call, including the colour mapping, and a colour mapping
splits the data into groups. So it fitted one line per group.

That inheritance is the useful default most of the time. When you do not want
it, give the layer its own `aes()` and it will use that one instead.

```{r}
#| label: fig-smooth-override
#| message: false
#| warning: false
#| fig-alt: |
#|   The same scatterplot, but with a single black trend line running across
#|   both clusters instead of one line per year.
#| fig-cap: "Overriding the inherited mapping for one layer only."
ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c
  )
) +
  geom_point(aes(colour = factor(year)), alpha = 0.4) +
  geom_smooth(method = "lm", colour = "black")
```

Now the colour mapping lives on the points, so the smooth doesn't see it, and
you get one line across everything.

Those two plots say quite different things, and the only difference is where the
`aes()` sits. Worth remembering when a smooth does something you didn't expect.

::: {.callout-note title="Your Turn"}

1. Add `geom_smooth()` to your humidity plot.
2. Swap the order of `geom_point()` and `geom_smooth()`. Can you see the
   difference? Look closely at where the lines and points overlap.
3. Try `geom_smooth(se = FALSE)`. What went away?

:::

## To summarise

Three things to take out of this chapter.

1. **A ggplot is data, aesthetics, and a geom, glued together with `+`.**
   Everything else is detail hanging off those three.
2. **Inside `aes()` maps a variable, outside `aes()` sets a value.** An
   unexpected legend means you mapped something by accident.
3. **The channels are not equal.** Position first, then colour, then shape, and
   size only when you have run out of better ideas.

And one habit, which is the one I actually care about.

When a plot surprises you, do not reach for a different example to copy. Ask
which of the seven pieces is doing something you did not ask it to do. Nearly
every ggplot2 problem I have ever had turned out to be one piece behaving
exactly as documented, in a way I had not thought about.

Next up is the shape of your data, which is where most of the remaining
surprises come from.


```{r}
#| echo: false
#| eval: false
# example nicer last plot
library(marquee)

# RColorBrewer::brewer.pal(name = "Dark2", n = 3) |> scales::show_col()
plot_cols <- RColorBrewer::brewer.pal(name = "Dark2", n = 3) |> 
  as.list() |>
  setNames(c("green", "orange", "purple"))

ggplot(
  oceanbuoys,
  aes(
    x = air_temp_c,
    y = sea_temp_c,
    colour = factor(year)
  )
) +
  geom_point(alpha = 0.6) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title = "Sea and air temperature from buoys in the tropical Pacific",
    x = "Air Temp (C)",
    y = "Sea Temp (C)",
    colour = "Year"
  ) +
  annotate(
    x = 24.5,
    y = 22.5,
    geom = "text",
    colour = plot_cols$green,
    size = 15,
    label = "1997",
    size.unit = "pt"
  ) +
  annotate(
    x = 26.5,
    y = 29.5,
    geom = "text",
    colour = plot_cols$orange,
    size = 15,
    label = "1993",
    size.unit = "pt"
  ) +
  theme(legend.position = "none",
        aspect.ratio = 1/1.618)
```

