# Making it land {#sec-land}

Your plot is right. Now it has to survive leaving your screen, and end up in a
paper, a slide, or a report where nobody's there to explain it.

This chapter is the last mile: labels, themes, combining plots, and getting a
file out that looks the way it did in RStudio.

We finish by turning the course into a checklist you can run on anything.

::: {.overview}

## Overview {.unnumbered}

**Duration** 41 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- When should I start polishing?
- What should the title say?
- How do I make every plot in a report match?
- How do I combine several plots?
- Why does the saved file look different to my screen?
- How do I critique a plot, including my own?

**What you need this session**

- A session of RStudio open
- The following packages installed

```{r}
#| label: pkgs
#| message: false
#| output: false
library(tidyverse)
library(naniar)
library(patchwork)
```

```{r}
#| label: pkgs-extra
#| echo: false
#| output: false
library(countdown)
```

```{r}
#| label: ch8-data
#| message: false
pedestrian_hourly <- pedestrian |>
        mutate(day_type = if_else(
                condition = str_starts(week_day, "S"),
                true = "weekend",
                false = "weekday"
                )
        ) |>
        group_by(sensor_name, day_type, hour) |>
        summarise(
                mean_count = mean(hourly_counts, na.rm = TRUE),
                .groups = "drop"
        )
```

:::

## Polish last

> **The idea:** polishing a plot you're going to throw away is the most common
> way to waste an afternoon.

Everything here comes **after** @sec-comparing, @sec-in-the-way and @sec-eye
have been answered. The order I work in:

1. `labs()`
2. colours with `scale_colour_*()`
3. a theme
4. move or remove the legend
5. `ggsave()`

Each step is cheaper to redo than the one before it.

Save the plot to an object first. Small habit, and from here everything is
`p + one thing`.

```{r}
#| label: fig-base-plot
#| fig-alt: |
#|   Four panels of hourly foot traffic, one per sensor, with weekday and
#|   weekend lines, and default axis labels.
p <- ggplot(pedestrian_hourly,
            aes(x = hour,
                y = mean_count,
                colour = day_type)) +
        geom_line() +
        facet_wrap(vars(sensor_name))

p
```

## Say the finding, not the variables

> **The idea:** the title is the loudest thing on the plot. Spend it on what you
> found, not on what the columns were called.

Default labels are your column names. `mean_count` is not a sentence.

```{r}
#| label: fig-labelled
#| fig-alt: |
#|   The same four panels with a title reading "Only the park gets busier on a
#|   weekend", a subtitle, and readable axis labels.
p_labelled <- p +
        labs(
                title = "Only the park gets busier on a weekend",
                subtitle = "Average hourly foot traffic, Melbourne, 2016",
                x = "Hour of the day",
                y = "People per hour",
                colour = NULL,
                caption = "Data: City of Melbourne, via the naniar package"
        )

p_labelled
```

Compare "Only the park gets busier on a weekend" with "mean_count by hour". One
of those tells you what you're looking at.

If you can't write that sentence, that's worth knowing **before** you spend an
hour on themes.

### Axes a human can read

The axis is text too, and the defaults are safe rather than readable.

- **Breaks.** ggplot2 picks `0 5 10 15 20` for hours, because it doesn't know
  they're hours. Nobody thinks in five hour blocks.
- **Big numbers.** Totals print as `2500000`.
- **Labels that collide.** Long category names overlap.

```{r}
#| label: fig-breaks
#| fig-alt: |
#|   The same plot with the x axis labelled midnight, 6am, noon and 6pm instead
#|   of 0, 5, 10, 15, 20.
p_labelled +
        scale_x_continuous(
                breaks = c(0, 6, 12, 18),
                labels = c("midnight", "6am", "noon", "6pm")
        )
```

For the other two:

```{r}
#| eval: false
# 2500000 becomes 2,500,000
scale_y_continuous(labels = scales::label_comma())

# drop labels that would collide, as the plot resizes
scale_x_discrete(guide = guide_axis(check.overlap = TRUE))

# or stagger them onto two rows
scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

`check.overlap` is the one worth knowing, because it thins labels automatically
when the same figure goes into a slide and a paper at different widths.

### Alt text

`labs(alt = )` sets what a screen reader says.

```{r}
#| label: alt-text
p_alt <- p_labelled +
        labs(alt = "Four line charts of average hourly foot traffic in
             Melbourne in 2016. Flagstaff Station and Spencer St have large
             weekday commuter peaks that vanish at the weekend. Bourke Street
             Mall is the same on both. Birrarung Marr is the only sensor
             busier at the weekend.")

get_alt_text(p_alt)
```

Say the finding, then the shape, then the axes. Not "a line chart of x against
y", which tells a screen reader user nothing they couldn't guess.

Every figure in this book has alt text. Look at the source of any of them.

::: {.callout-note title="Your Turn"}

1. Write a title for a plot of your own that states its finding.

2. Add alt text and read it back:

```{r}
#| eval: false
my_plot <- my_plot +
        labs(alt = "___")

get_alt_text(my_plot)
```

3. Put comma separators on a y axis running into the millions.

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** If you can't write it, the plot probably isn't finished. That's the
useful outcome, not a failure.

**3.** `scale_y_continuous(labels = scales::label_comma())`.

:::

:::

### What to hold onto

- The title is the loudest element. Spend it on the finding
- If you can't state the finding, stop polishing and go back to the plot
- Alt text says the finding first, not the chart type

## Themes, and making one yours

> **The idea:** a theme is every non-data decision in one place. Wrap it in a
> function and every plot in the report matches.

:::{.panel-tabset}

## minimal

```{r}
#| label: fig-theme-minimal
#| fig-alt: The labelled plot with theme_minimal applied.
p_labelled + theme_minimal()
```

## bw

```{r}
#| label: fig-theme-bw
#| fig-alt: The labelled plot with theme_bw applied.
p_labelled + theme_bw()
```

## classic

```{r}
#| label: fig-theme-classic
#| fig-alt: The labelled plot with theme_classic applied.
p_labelled + theme_classic()
```

## light

```{r}
#| label: fig-theme-light
#| fig-alt: The labelled plot with theme_light applied.
p_labelled + theme_light()
```

:::

Then `theme()` for the details, and wrap the lot in a function:

```{r}
#| label: fig-theme-fn
#| fig-alt: |
#|   The labelled plot with a custom theme: minimal base, legend at the bottom,
#|   no minor gridlines, title aligned to the left edge.
theme_pedestrian <- function(...) {
        theme_minimal(...) +
                theme(
                        legend.position = "bottom",
                        plot.title.position = "plot",
                        panel.grid.minor = element_blank()
                )
}

p_labelled + theme_pedestrian()
```

Change the function once and every figure in the paper changes.

`theme_set(theme_pedestrian())` at the top of a document does it for everything
without you typing it each time.

::: {.callout-tip title="Going deeper: other people's themes" collapse="true"}

`ggthemes` and `hrbrthemes` ship dozens of ready made themes, including
deliberate imitations of the Economist, FiveThirtyEight and the Wall Street
Journal.

They're fun and worth ten minutes of play. The skill transfers either way,
because they're all just `theme()` calls somebody else wrote.

One catch: `hrbrthemes` wants specific fonts installed and complains loudly when
they're missing.

:::

::: {.callout-note title="Your Turn"}

1. Write your own theme function. Change one thing you always change.

```{r}
#| eval: false
theme_mine <- function(...) {
        theme_minimal(...) +
                theme(___)
}
```

2. Try `theme_minimal(base_size = 16)`. Where does the `...` in your function
   send that?

3. Run `theme_set(theme_mine())`, then draw any plot. What happened?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**2.** Straight through to `theme_minimal()`, which is why the `...` is there.
Without it your function would silently ignore every argument.

**3.** Every plot afterwards uses it, with no `+ theme_mine()` needed. Handy for
a report, confusing in a script somebody else reads.

:::

:::

### What to hold onto

- A theme is every non-data decision in one place
- Wrapping it in a function is what makes a report look like one document
- `theme_set()` when you don't want to type it at all

## Combining plots

> **The idea:** patchwork combines finished plots with the operators you'd
> guess.

```{r}
#| label: fig-patchwork
#| fig-alt: |
#|   Two line charts side by side, one for weekdays and one for weekends, each
#|   with four coloured sensor lines.
p_weekday <- pedestrian_hourly |>
        filter(day_type == "weekday") |>
        ggplot(aes(x = hour, y = mean_count, colour = sensor_name)) +
        geom_line() +
        labs(title = "Weekday")

p_weekend <- pedestrian_hourly |>
        filter(day_type == "weekend") |>
        ggplot(aes(x = hour, y = mean_count, colour = sensor_name)) +
        geom_line() +
        labs(title = "Weekend")

p_weekday + p_weekend
```

- `p1 + p2` side by side
- `p1 / p2` stacked
- `(p1 + p2) / p3` nested

Two legends saying the same thing is the decoration @sec-in-the-way was about.
`plot_layout(guides = "collect")` fixes it.

```{r}
#| label: fig-patchwork-collect
#| fig-alt: |
#|   The same two charts with a single shared legend underneath and panel tags
#|   A and B.
p_weekday + p_weekend +
        plot_layout(guides = "collect") +
        plot_annotation(
                title = "Only the park gets busier on a weekend",
                tag_levels = "A"
        ) &
        theme(legend.position = "bottom")
```

`tag_levels = "A"` labels the panels, which is what a journal will ask for.

::: {.callout-note title="Your Turn"}

1. Build `(p_weekday + p_weekend) / p`. What do the brackets do?

2. Collect the legends and move them to the right instead of the bottom.

3. Rebuild the four sensor story from @sec-comparing as one figure.

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** The brackets group the two plots so they share a row, and the third sits
underneath across the full width. Without them you get three in a row.

**2.** Change `legend.position` to `"right"` in the `&` line. The `&` applies
the theme to every plot in the patchwork, rather than just the last one, which
is the bit that catches people.

:::

:::

### What to hold onto

- `+` for side by side, `/` for stacked, brackets to group
- `plot_layout(guides = "collect")` for one legend instead of three
- `&` applies a theme to the whole patchwork, `+` only to the last plot

## Getting it out

> **The idea:** the saved file does not look like the RStudio pane, and knowing
> why saves you an afternoon.

```{r}
#| eval: false
ggsave(
        filename = "pedestrian-weekend.png",
        plot = p_labelled,
        width = 8,
        height = 5,
        dpi = 300
)
```

Text does not scale with the plot. Make the canvas bigger and the text stays the
same physical size, so it gets proportionally smaller.

```{r}
#| eval: false
# same plot, same dpi, same code
ggsave("narrow.png", p_labelled, width = 4, height = 3)
ggsave("wide.png", p_labelled, width = 12, height = 9)
```

Open both. The wide one has tiny text and nothing about the code changed.

Things worth knowing:

- `ggsave()` defaults to the **last plot drawn**, which bites once you've moved
  on. Pass `plot =` explicitly
- `width` and `height` are inches unless you set `units =`
- `dpi = 300` for print, and it does nothing for a vector format
- the file extension picks the device
- vector (`.pdf`, `.svg`) for line art, raster (`.png`) for anything with
  thousands of overplotted points. A hex plot from @sec-in-the-way is small as a
  PDF; 35,152 points is not

::: {.callout-note title="Your Turn"}

1. Save the same plot at `width = 4` and `width = 12`. Open both. What happened
   to the text?

2. Save one as `.png` and one as `.pdf`. Compare the file sizes.

3. You need a figure for a journal: one column wide, 300 dpi. What do you pass?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**3.** Something like `ggsave("fig1.pdf", plot = p, width = 3.5, height = 3,
dpi = 300)`. One column is usually about 3.5 inches. Check the journal's guide,
they all publish one.

:::

:::

### What to hold onto

- Text doesn't scale with the canvas, so size the canvas to its final home
- Always pass `plot =`
- Vector for line art, raster for dense point clouds

## Four questions

The chapter titles of this course are a checklist. That was deliberate.

> **What am I comparing?** (@sec-comparing)
> Is the thing I want compared next to the thing I want it compared to?
>
> **What's in the way?** (@sec-in-the-way)
> How much of this is not data, and can I see the data through it?
>
> **Where should the eye go?** (@sec-eye)
> What does a reader meet first, and did I choose it?
>
> **Does it say its finding?** (this chapter)
> If somebody reads only the title, do they get the point?

And a fifth, quieter one we've been doing since @sec-anatomy:

> **Did I draw it first?**

::: {.callout-note title="Your Turn"}

Run the four questions on a plot you've made before this course.

Then run them on a plot from this book. Mine included: there are plenty in here
that would not survive all four.

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 10)
```

:::

## Open practice and questions

Bring a plot you're working on, or pick something from the [ggplot2 extension
gallery](https://exts.ggplot2.tidyverse.org/gallery/) you'd like to try.
