# The shape of your data {#sec-shape}

<!-- Claude: "shape" is the wrong word here - it should be "tidy data" - the key is "tidy". Make sure to update other references to "shape" here. -->

In my experience, a lot of my ggplot problems were actually problems with my 
data, and how I had it formatted. I found this was especially so early on, when 
I was learning how to use ggplot.

I believe these data problems are somewhat solved by making sure your data 
is in a format called "tidy data". We will talk about what that means, and 
how to spot data that isn't "tidy", and fix it.

We also come back to those missing observations in the oceanbuoys data we look at in @sec-anatomy.

::: {.overview}

## Overview {.unnumbered}

**Duration** 40 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What does ggplot2 expect my data to look like?
- How do I make sure the spreadsheet was actually given is "tidy data"?
- Why did my line graph come out as a scribble?
- What happened to the rows with missing values?

:::

```{r}
#| label: pkgs
library(dplyr)
library(tidyr)
```


## What ggplot2 expects: Tidy Data

The `ggplot2` package expects data in a specific format, called "Tidy Data".
Tidy Data is defined as:

1. Every variable has it's own column
2. Every observation must have it's own row
3. Each cell must have its own cell

Let's explore this definition by breaking some of the rules. We will look again at the `pen_short` data:

```{r}
#| echo: false
pen_short <- penguins |> 
        as_tibble() |> 
        select(species, year, bill_len, bill_dep) |> 
        drop_na() |> 
        group_by(species, year) |> 
        slice(1) |> 
        ungroup()
```

This is the data in **tidy data form**

```{r}
pen_short
```

Think back to the plot of `pen_short` that we had in the previous chapter:

```{r}
library(ggplot2)
ggplot(pen_short,
       aes(x = bill_len,
           y = bill_dep,
           colour = species)) + 
        geom_point(size = 5)
```

Notice how each of the columns map onto an argument inside `aes()` ?

Here it is as messy - notice the variables don't form columns - they are in the rows:

```{r}
#| echo: false
pen_long_bill <- pen_short |> 
        pivot_longer(
                cols = c(bill_len, bill_dep),
                names_to = "bill_variable",
                values_to = "measurement"
        )
pen_long_bill
```

Here it is as another flavour of messy - each individual cell contains two values:

```{r}
#| echo: false
pen_two_val <- pen_short |> 
        mutate(bill_ratio = paste0(bill_len,"/",bill_dep)) |> 
        select(-bill_len,
               -bill_dep)

pen_two_val
```

And yet another messy - this time the year variable is for each column. These each contain measurements of bill depth, and bill length, respectively:

```{r}
#| echo: false
# bill_dep
pen_year_bill_dep <- pen_short |> 
        select(-bill_len) |> 
        pivot_wider(
                names_from = year,
                values_from = bill_dep
        )

pen_year_bill_dep

# bill_len
pen_year_bill_len <- pen_short |> 
        select(-bill_dep) |> 
        pivot_wider(
                names_from = year,
                values_from = bill_len
        )

pen_year_bill_len
```

Now let's go back to the tidy data of penguins:

```{r}
pen_short
```

And go back again to our ggplot:

```{r}
library(ggplot2)
ggplot(pen_short,
       aes(x = bill_len,
           y = bill_dep,
           colour = species)) + 
        geom_point(size = 5)
```

Before, each of the columns map onto an argument inside `aes()`.

Now, think about how to do this for a messy dataset:

```{r}
pen_long_bill
```

How can we plot "bill depth vs bill length"?

And when there are two values in each cell, how do we plot those?

```{r}
pen_two_val
```

And for the separate years? What do we call these?

```{r}
pen_year_bill_len
```

I'm not saying you can't plot data from this data. I am saying it is hard to do it. It doesn't fit into the tool usage.

But when the data is in a standard, "Tidy" form, then this becomes much easier:

```{r}
pen_short
```


I am not sure if I can overstate the impact of the concept of "Tidy Data" on
the R programming world, and statistics in general. 

## Making data tidy

To make this a bit easier to learn, we have our reference set of tidy data, that each of these will convert back to:

```{r}
pen_short
```

To do the data tidying, we will be looking at two functions:

1. `pivot_wider()`: To make our data have more columns, and fewer rows
2. `pivot_longer()`: To make our data have fewer columns, and more rows.

The problem with `pen_long_bill` here is that all the "bill" related variables are in a single column. 

```{r}
pen_long_bill
```

We want to have one variable per column, so we want: "bill_variable" and "measurement" turned into two variables:

- "bill_len"
- "bill_dep"

Where their observations form rows - the values of bill length, and bill depth, respectively.

To make this happen we use `pivot_wider()`, and use the `names_from` and `values_from` arguments:

```{r}
pen_long_bill |> 
        pivot_wider(
                names_from = bill_variable,
                values_from = measurement
        )
```

- `names_from`: Here we want to create more columns, to make them wider, and we want the **names** of those columns to come **from** `bill_variable`
- `values_from`: We need the **values** for our new columns to come **from** `measurement` 

The way I internalise this when I am using the `pivot_wider()` or `pivot_longer()` functions, is to always think of:

- Names: What are my tidy columns
- Values: What are my values


::: {.callout-note title="Your Turn"}

Run the code below to create the `hi_score` dataset:

```{r}
hi_score <- tibble::tribble(
  ~score, ~variable, ~values,
  11000,  "game",    "pacman",
  11000,  "person",  "James",
  9100,   "game",    "asteroids",
  9100,   "person",  "James",
  18000,  "game",    "pacman",
  18000,  "person",  "Sarida",
  1200,   "game",    "asteroids",
  1200,   "person",  "Sarida",
)

hi_score
```

Now, tidy this using `pivot_wider()`

<details>
<summary>Answer code below</summary>

<details>
<summary>Only really click on this if you've actually had a go at doing it!</summary>

```{r}
hi_score |> 
        pivot_wider(
                names_from = variable,
                values_from = values
        ) |> 
        # use dplyr::relocate
        relocate(
                person, game, score
        )
```

</details>

</details>

Now try using `pivot_wider()` with a different dataset: `fish_encounters` - which contains (type `?fish_encounters` to get details) information about fish swimming down a river.

```{r}
fish_encounters
```

Tidy this using `pivot_wider()`:

<details>
<summary>Answer code below</summary>

<details>
<summary>Only really click on this if you've actually had a go at doing it!</summary>

```{r}
fish_encounters |> 
        pivot_wider(
                names_from = station,
                values_from = seen
        )
```


</details>
</details>

What do you notice about this data? If there are missing values, should we do something about them? See the argument `values_fill` for options here.

<details>
<summary>Answer code below</summary>

```{r}
fish_encounters |> 
        pivot_wider(
                names_from = station,
                values_from = seen,
                values_fill = 0
        )
```

</details>

:::

## Using `pivot_longer()`

Let's look at using `pivot_longer()` with `pen_year_bill_dep`

```{r}
pen_year_bill_dep
```

Similar to `pivot_wider()`, there `name` and `value` arguments - this time, we have `names_to`, `values_to`, and `cols`:


```{r}
pen_year_bill_dep |> 
        pivot_longer(
                cols = c(`2007`, `2008`, `2009`),
                names_to = "year",
                values_to = "bill_dep"
        )
```

- `cols`: We use `cols` as we need to tell `pivot_longer()` which columns to be acting on. We didn't need to do that with `pivot_wider()` as we told it already which columns were being acted on. Note that we need to wrap the years in back ticks (``), as they start with numbers. See below for a workaround.
- `names_to`: We specify "years" in quotes as this is the name of the column we are creating
- `values_to`: we specify "bill_dep"

It is worth explaining some workarounds for `cols` - you don't always want to be specifying every single column name.  We can be a little bit crafty and use `-species`, which says: every column but species.

```{r}
pen_year_bill_dep |> 
        pivot_longer(
                cols = -species,
                names_to = "year",
                values_to = "bill_dep"
        )
```

We can also do `` `2007`:`2009` `` to indicate all columns between 2007 and 2009. 

```{r}
pen_year_bill_dep |> 
        pivot_longer(
                cols = `2007`:`2009`,
                names_to = "year",
                values_to = "bill_dep"
        )
```

Note that you cannot do this with just the bare numbers, as it will assume you are referring to column positions by number, not name:

```{r}
#| error: true
pen_year_bill_dep |> 
        pivot_longer(
                cols = c(2007, 2008, 2009),
                names_to = "year",
                values_to = "bill_dep"
        )
```

See `help("tidyr_tidy_select")` for more information on specifying column names.

::: {.callout-note title="Your Turn"}

Now try using `pivot_longer()` with a different dataset: `relig_income`, which comes bundled inside `tidyr`. This is a dataset that contains religion and income ranges.

```{r}
relig_income
```

:::

## Separating columns to make them tidy

When values are combined together, you can use the `separate_wider_delim()` function to separate them out: 

```{r}
pen_two_val |> 
        separate_wider_delim(
                cols = bill_ratio,
                delim = "/",
                names = c("bill_dep", "bill_len")
                )

```


::: {.callout-note title="Your Turn"}

Now try with ...

:::

<!-- The same pedestrian data with the four sensors as four columns, which is
     exactly how it arrives from anyone's Excel file. pivot_longer() to fix it,
     pivot_wider() as the way back. -->


## The scribble

<!-- geom_line() across four sensors with no group. Comes out as a solid black
     mess. Both `group` and `colour` fix it, and the difference between them is
     worth two minutes. -->

::: {.callout-note title="Your Turn"}

1. TODO: exercise for this section.

:::

## Missing Data

<!-- Back to oceanbuoys and the `Removed 81 rows` warning from @sec-anatomy.
     vis_miss() and geom_miss_point(), which draws the dropped rows instead of
     discarding them.

     It belongs in this chapter because missingness is a property of the data
     you have, not of what the plot computes. And the payoff lands on the same
     data that raised the question. -->

::: {.callout-note title="Your Turn"}

1. TODO: exercise for this section.

:::

## Extension exercise

Now, look at the `us_rent_income` dataset

```{r}
us_rent_income
```
