# What's in the way? {#sec-in-the-way}

You've got the right comparison on the page and it still looks like a mess. Now
what?

Three things get between a reader and your data: Decoration, density, and
summary.

This chapter takes them one at a time, and most of the work is deciding what to
remove.

::: {.overview}

## Overview {.unnumbered}

**Duration** 49 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- How much of this plot is not data?
- What do I do when there are too many points to see anything?
- What is my summary hiding?

**What you need this session**

- A session of RStudio open
- Something to draw on, and something to draw with
- The following packages installed

```{r}
#| label: pkgs
#| message: false
#| output: false
library(tidyverse)
library(naniar)
library(gghighlight)
library(ggridges)
library(ggrain)
```

```{r}
#| label: pkgs-extra
#| echo: false
#| output: false
library(countdown)
```

```{r}
#| label: ch6-data
#| message: false
pedestrian_hourly <- pedestrian |>
        mutate(day_type = if_else(
                condition = str_starts(week_day, "S"),
                true = "weekend",
                false = "weekday"
                )
        ) |>
        group_by(sensor_name, day_type, hour) |>
        summarise(
                mean_count = mean(hourly_counts, na.rm = TRUE),
                .groups = "drop"
        )
```

:::

## Decoration

> **The idea:** every plot has ink that is data and ink that is not. The
> not-data ink still costs the reader a look.

Here is where we left off in @sec-comparing.

```{r}
#| label: fig-before
#| fig-alt: |
#|   Four panels, one per sensor, each with a weekday and a weekend line, on
#|   ggplot2's default grey background with white gridlines and a legend on the
#|   right.
p_busy <- ggplot(pedestrian_hourly,
                 aes(x = hour,
                     y = mean_count,
                     colour = day_type)) +
        geom_line() +
        facet_wrap(vars(sensor_name))

p_busy
```

Eight lines are data. Not data:

- the grey panel background
- the white gridlines, major and minor
- the box around the legend
- the legend title, which is a column name

Edward Tufte called the useful fraction the **data-ink ratio**. Go through each
piece and ask: is this helping, or is it just here?

:::{.panel-tabset}

## default

```{r}
#| label: fig-strip-0
#| fig-alt: The four panel plot with ggplot2's default grey theme.
p_busy
```

## theme_minimal()

```{r}
#| label: fig-strip-1
#| fig-alt: The same plot with a white background and light grey gridlines.
p_busy + theme_minimal()
```

## fewer gridlines

```{r}
#| label: fig-strip-2
#| fig-alt: The same plot again with the minor gridlines removed.
p_busy +
        theme_minimal() +
        theme(panel.grid.minor = element_blank())
```

## legend below

```{r}
#| label: fig-strip-3
#| fig-alt: |
#|   The same plot with the legend moved underneath, giving the panels more
#|   width.
p_busy +
        theme_minimal() +
        theme(panel.grid.minor = element_blank(),
              legend.position = "bottom")
```

:::

Each step removed something and nothing was lost. The last one made the panels
wider too, because the legend was using horizontal space to say two words.

### Colour is ink

Colour counts as ink, and it is the piece people most often spend twice.

```{r}
#| label: fig-colour-redundant
#| warning: false
#| fig-alt: |
#|   Four boxplots, each filled with a different colour, with a legend on the
#|   right repeating the sensor names already on the x axis.
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts,
           fill = sensor_name)) +
        geom_boxplot()
```

The x axis already says which sensor is which. The fill says it again, and then
a legend says it a third time.

```{r}
#| label: fig-colour-dropped
#| warning: false
#| fig-alt: |
#|   The same four boxplots in plain grey, with no legend, and the sensor names
#|   still labelled on the x axis.
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts)) +
        geom_boxplot()
```

Same message, less ink, and the plot got wider.

Colour earns its place when it carries something no other channel is carrying.
If dropping it costs you nothing, it was decoration.

::: {.callout-tip title="Going deeper: data-ink is a direction, not a target" collapse="true"}

Push this to its end and you get a plot with no gridlines, no axis and no
legend, which is unreadable.

Gridlines are the case worth arguing about. They are not data, and they are how
a reader gets a number off an axis. Delete them and you've made the plot cleaner
and less useful.

So the question is never "how much can I remove". It's "what is this piece for",
and if you can answer that, it stays.

:::

::: {.callout-note title="Your Turn"}

1. Run `p_busy + theme_void()`. What did you lose, and would you ever want that?

2. Try each of these and pick one you'd keep:

```{r}
#| eval: false
p_busy + theme_bw()
p_busy + theme_classic()
p_busy + theme_light()
```

3. The legend on `p_busy` is titled `day_type`. Fix it.

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** `theme_void()` removes the axes too, so you can no longer read a value off
the plot. Useful for maps, and for sparklines where the shape is the whole
message. Rarely anything else.

**3.** `labs(colour = NULL)` drops the title, since "weekday" and "weekend"
explain themselves. `labs(colour = "Day type")` if you'd rather keep it.

:::

:::

### What to hold onto

- Removing non-data ink is cheap and usually helps
- Colour is ink. If dropping it costs nothing, it was decoration
- Data-ink is a direction, not a score. Gridlines usually have a job

## Density

> **The idea:** when there are more points than pixels, the plot stops being a
> picture of your data and becomes a black shape.

Everything above was 192 summarised rows. Here is the raw data, all 37,700.

```{r}
#| label: fig-overplot
#| warning: false
#| fig-alt: |
#|   A scatterplot of pedestrian count against hour, drawn as solid black
#|   columns of points with no visible structure.
ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_point()
```

There could be ten points in the middle of that or ten thousand.

Three moves, in the order I try them:

- **`alpha`**, so overlapping points get darker
- **`geom_jitter()`**, to spread points that share an x value
- **a geom that bins**, once there are too many for either to help

```{r}
#| label: fig-alpha
#| warning: false
#| fig-alt: |
#|   The same scatterplot with very transparent points, so denser regions
#|   appear darker and the daily shape emerges.
ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_point(alpha = 0.05)
```

Now you can see most hours are quiet and the busy ones are rare.

`hour` is a whole number, so every point in a column sits at the same x and we
are still stacking them. `geom_jitter()` spreads them sideways.

```{r}
#| label: fig-jitter
#| warning: false
#| fig-alt: |
#|   The same data jittered horizontally, so each hour becomes a cloud rather
#|   than a line of points.
ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_jitter(alpha = 0.05, width = 0.3)
```

::: {.callout-important title="This will bite you: jitter moves your data"}

A jittered point is in the wrong place on purpose.

Fine when you're reading density. Not fine when somebody reads a value off the
axis, because at `width = 0.3` the 8am readings sit over 7:42 and 8:18.

:::

`geom_hex()` bins in two dimensions and colours each bin by how many landed in
it. It's @sec-computed's binning, with one more axis.

```{r}
#| label: fig-hex
#| warning: false
#| fig-alt: |
#|   A hexagonal binned plot of the same data, with colour showing how many
#|   readings fall in each hexagon.
ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_hex(bins = 30)
```

Density is now something you read off a legend, not guess from how black the
page looks.

Same trade as `geom_rug()` in @sec-computed: a rug worked on 397 points and
would be a smear on 35,152.

The other kind of crowding is too many **series**. `gghighlight()` keeps one and
greys the rest.

```{r}
#| label: fig-gghighlight
#| warning: false
#| message: false
#| fig-alt: |
#|   Four sensor lines with Flagstaff Station in colour and the other three
#|   greyed out behind it.
ggplot(pedestrian_hourly,
       aes(x = hour,
           y = mean_count,
           colour = sensor_name)) +
        geom_line() +
        gghighlight(sensor_name == "Flagstaff Station")
```

::: {.callout-note title="Your Turn"}

1. Try `alpha = 0.01`, then `alpha = 0.3`. Which lets you see the shape?

2. Swap `geom_hex()` for `geom_bin2d()`. What changed?

3. Change the bins:

```{r}
#| eval: false
ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_hex(bins = ___)
```

   Try 10, then 60. Where have you met this decision before?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** At `0.01` almost everything vanishes. At `0.3` the busy hours saturate to
solid black and you're back where you started. Around `0.05` works here, and the
right number depends on how many points you have.

**2.** Squares instead of hexagons, same computation. Hexagons don't line up
into rows and columns, so your eye is less likely to invent a grid.

**3.** It's the binwidth decision from @sec-computed. Too few and the shape goes,
too many and you're counting individual points again.

:::

:::

### What to hold onto

- `alpha` first, it's free
- `geom_jitter()` for ties, but it moves your data
- `geom_hex()` once there are more points than pixels
- `gghighlight()` when the crowding is series, not points

## Summary

> **The idea:** a summary is a claim about the shape of your data. When the
> shape is wrong, the summary is confidently wrong.

Four sensors as boxplots.

```{r}
#| label: fig-box-sensors
#| warning: false
#| fig-alt: |
#|   Four boxplots, one per sensor. Each box is squashed near the bottom of the
#|   range with a long column of outlier dots above it.
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts)) +
        geom_boxplot()
```

Look at the dots above the whiskers. ggplot2 is calling those outliers:

```{r}
#| label: outlier-count
pedestrian |>
        group_by(sensor_name) |>
        filter(hourly_counts >
                       quantile(hourly_counts, 0.75, na.rm = TRUE) +
                       1.5 * IQR(hourly_counts, na.rm = TRUE)) |>
        count(sensor_name)
```

Flagstaff has 663. They're the commute, which is the reason the sensor is there.

We know why from @sec-computed: 8am at Flagstaff is two mornings, not one. A
boxplot assumes one lump. When there are two, it describes a middle that nothing
is near and flags the real story as anomalous.

### Show the data on top of the summary

A violin keeps the shape. Layering the points on top keeps the evidence.

```{r}
#| label: fig-violin-points
#| warning: false
#| fig-alt: |
#|   Four violin plots with the raw readings jittered over the top, so both the
#|   estimated shape and the actual points are visible.
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts)) +
        geom_violin() +
        geom_jitter(alpha = 0.02, width = 0.15)
```

Two layers, and now the summary has its working shown. Flagstaff's second
population is a bulge in the violin *and* a band of real points.

`geom_rain()` does the same job with a boxplot, a half violin and the points.

```{r}
#| label: fig-rain
#| warning: false
#| message: false
#| fig-alt: |
#|   A raincloud plot for each sensor: a half violin, a boxplot, and the raw
#|   points scattered below.
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts)) +
        geom_rain(alpha = 0.02)
```

### Ridgelines

When you have a lot of groups, stack the distributions instead of standing them
up side by side.

```{r}
#| label: fig-ridge-hour
#| message: false
#| warning: false
#| fig-alt: |
#|   Twenty four density curves stacked vertically, one per hour of the day at
#|   Flagstaff Station, showing the two commuter peaks emerging and fading.
pedestrian |>
        filter(sensor_name == "Flagstaff Station") |>
        ggplot(aes(x = hourly_counts,
                   y = factor(hour))) +
        geom_density_ridges()
```

Twenty four distributions, one per hour, and you can read the day down the page.
The two-humped hours are visible as two-humped curves.

Try that as twenty four boxplots and you get twenty four squashed boxes.

::: {.callout-tip title="Read more"}

The classic demonstration that a summary can be confidently wrong is Anscombe's
quartet: four datasets with identical means, variances, correlations and
regression lines, and four different pictures.

I worked it through the tidyverse here: [Tidyverse Case Study: Anscombe's
quartet](https://www.njtierney.com/posts/2020-06-01-tidy-anscombe/).

The punchline is in the base R helpfile's own comment:

> `## Now, do what you should have done in the first place: PLOTS`

:::

::: {.callout-tip title="Going deeper: comparing a panel to everything" collapse="true"}

Faceting answers "what does this place look like". It can't answer "compared to
what", because each panel only holds its own data.

Give one layer its own `data =` with the faceting column removed, and it appears
in every panel:

```{r}
#| label: fig-background
#| warning: false
#| fig-alt: |
#|   Four panels, each showing all the data in light grey with that panel's own
#|   sensor drawn over the top in blue.
pedestrian_background <- pedestrian |>
        select(-sensor_name)

ggplot(pedestrian,
       aes(x = hour,
           y = hourly_counts)) +
        geom_point(data = pedestrian_background,
                   colour = "grey80",
                   alpha = 0.05) +
        geom_point(alpha = 0.05,
                   colour = "steelblue") +
        facet_wrap(vars(sensor_name))
```

Every panel now shows its own sensor against the shape of all four.

:::

::: {.callout-note title="Your Turn"}

1. Sketch a violin of **Birrarung Marr**, given a park doesn't care what day it
   is. Then draw it.

2. Layer a boxplot inside the violin:

```{r}
#| eval: false
ggplot(pedestrian,
       aes(x = sensor_name,
           y = hourly_counts)) +
        geom_violin() +
        geom_boxplot(width = ___)
```

3. Make a ridgeline of the four sensors instead of the hours. Which grouping
   suits ridgelines better?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** One lump, low and wide. No commute means no second population, so there's
nothing for the boxplot to hide.

**2.** `width = 0.1`. You get the median and quartiles as numbers you can read,
sitting inside the shape that tells you whether those numbers mean anything.

**3.** Hours, because there are 24 of them and they have an order. Ridgelines
work when groups are many and sequential. With four unordered sensors, side by
side violins are easier to compare.

:::

:::

### What to hold onto

- A boxplot claims your data has one middle
- When that's false it hides the finding and calls it an outlier
- Layer the points over the summary when you can afford the ink
- Ridgelines for many ordered groups, violins for a few

## To summarise

Three things to take out of this chapter.

1. **Non-data ink is not free**, and colour is ink. If dropping it costs
   nothing, it was decoration.
2. **Too many points is a geom problem.** `alpha` buys a little, `geom_hex()`
   solves it by counting.
3. **A summary is a claim about shape.** Show the data on top of it and the
   claim has to hold up.

When a plot looks like a mess, don't start styling it. Ask what is between the
reader and the data, and take that out first.

Next up is where the eye goes: not what to take away, but what to make loud.
