# What gets computed {#sec-computed}

Some geoms draw exactly what you expect: `geom_point()`, `geom_line()`.

Others need to do some computation! `geom_histogram()`, `geom_density()`, `geom_bar()`, and `geom_boxplot()` are all examples of this.

This chapter is what happens between your data and the vis on screen.

::: {.overview}

## Overview {.unnumbered}

**Duration** 40 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What is a boxplot actually showing me, and what is it hiding?
- When do I want `geom_bar()`, and when do I want `geom_col()`?
- How do I choose a binwidth, and how do I show the raw data too?

**What you need this session**

- A session of RStudio open
- Something to draw on, and something to draw with
- The following packages installed

```{r}
#| label: pkgs
#| message: false
#| output: false
library(dplyr)
library(ggplot2)
library(naniar)
```

```{r}
#| label: pkgs-extra
#| echo: false
#| output: false
library(countdown)
```

:::

## New data: Melbourne pedestrians

We are changing datasets for this chapter - looking at how people move in a city! The City of Melbourne has sensors that count how many people
walk past, every hour. A year of that data ships with {naniar}:

```{r}
#| label: ped-glimpse
pedestrian
```

Each row is the number of counts of pedestrians in a one hour period, at a given sensor, for every hour of 2016. There are four sensors.

```{r}
#| label: ped-sensors
pedestrian |>
  distinct(sensor_name)
```

These are quite different places in Melbourne city:

- **Flagstaff Station** a train station
- **Birrarung Marr** parkland along the river
- **Bourke Street Mall** mostly shopping
- **Spencer St-Collins St** a corner in the middle of the CBD 

This gives us some interesting data to play with.

## Geoms that summarise for you

> **The idea:** a geom can take all your raw numbers and hand back a few. That
> is useful, and it means the plot is showing you a decision somebody made, not
> the data itself.

Let's start with a question: 

> What does a day at Flagstaff Station look like?

Flagstaff is a train station, so before we write anything, have a think about
what that should look like. Here's a snippet of one day of data for Flagstaff:

```{r}
#| label: flagstaff-data
flagstaff <- pedestrian |>
  filter(sensor_name == "Flagstaff Station")

flagstaff_january <- flagstaff |> 
  filter(
    month == "January"
  )

flagstaff_january_1 <- flagstaff_january |> 
  filter(
    month_day == 1
  )

flagstaff_january_1 |> 
  knitr::kable()
```


::: {.callout-tip title="Starting from here? Run this first" collapse="true"}

```{r}
#| ref.label: ["pkgs"]
#| eval: false
#| echo: true
```

:::

::: {.callout-note title="Your Turn: draw it first"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs"]
#| eval: false
#| echo: true
```

:::
Take your pen and paper, and draw the chart you would like to see to explore:

- the number of people (hourly_counts)
- walking by over time (hour)

You don't need exact number on the axes. I am just curious on the **shape**. You can use bars, dots, lines, other summaries, whatever you want!

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```


- Now, how do you think that chart might change if we focus on:
  - A given time - like 8am
  - A given day - like Monday, Friday, Sunday?

How much would they vary? How would you show how they compare?

:::

We have a reading for each hour of every day, so nearly 400 readings for each
hour of the day. A geom that takes all of them
and draws the spread is `geom_boxplot()`.

Let's look at just flagstaff:

```{r}
flagstaff
```


And plot it as a boxplot

```{r}
#| label: fig-box-nogroup
#| warning: false
#| fig-alt: |
#|   A single wide boxplot spanning the whole width of the panel, rather than
#|   one box per hour.
#| fig-cap: "One box. Not what I wanted."
ggplot(flagstaff,
       aes(x = hour,
           y = hourly_counts)) +
  geom_boxplot()
```

Hmmm - we get a single box - this isn't what we want.

We saw something similar earlier - in @sec-shape, `geom_line()` sawtoothed across the whole plot, because it did not know species was a thing. We even get a few (!) warnings from ggplot2:

```
Warning messages:
1: Orientation is not uniquely specified when both the x and y aesthetics are continuous. Picking default
orientation 'x'. 
2: Continuous x aesthetic
ℹ did you forget `aes(group = ...)`? 
3: Removed 1 row containing non-finite outside the scale range (`stat_boxplot()`).
```

The one we are interested in here is:

```
2: Continuous x aesthetic
ℹ did you forget `aes(group = ...)`? 
```

`hour` is a number, so ggplot2 treated it as continuous - not something that is a whole number/integer. To get ggplot to treat it as a group, use `group`:

```{r}
#| label: fig-box-hour
#| warning: false
#| fig-alt: |
#|   Twenty four boxplots, one per hour. Very low flat boxes overnight, a large
#|   box at 8am spanning roughly 200 to 5100, a moderate bump around midday, and
#|   a second large box around 5pm.
#| fig-cap: "A day at Flagstaff Station, with the variation left in."
ggplot(flagstaff,
       aes(x = hour,
           y = hourly_counts,
           group = hour)) +
  geom_boxplot()
```

How does this compare to your drawing.

The commuter shape is there, with a peak at 8am and another around 5pm. But the
thing I want you to notice is the **8am box** - it has a big range!

### What did it compute?

A boxplot is five numbers, worked out for you from the raw data. We typically call this the "five number summary": min, quartile 1 (25%), median, quartile 3 (75%), max.

::: {.callout-caution title="Some history: where the boxplot came from" collapse="true"}

The boxplot is John Tukey's, from his 1977 book *Exploratory Data Analysis*, and
it was designed to be drawn **by hand**, quickly, on paper. That is why it is
five numbers and not fifty: five is what you can work out and draw in a minute.

Tukey called the box edges **hinges** rather than quartiles. They are almost the
same thing and not quite, which is why `?geom_boxplot` and `boxplot.stats` both
hedge slightly about how the box is computed.

He is also the reason we say "bit" and "software". He coined both.

:::

```{r}
#| label: five-number

flagstaff_8am <- flagstaff |>
  filter(hour == 8)

flagstaff_8am |> 
  summarise(
    min = min(hourly_counts, na.rm = TRUE),
    q1 = quantile(hourly_counts, 0.25, na.rm = TRUE),
    median = median(hourly_counts, na.rm = TRUE),
    q3 = quantile(hourly_counts, 0.75, na.rm = TRUE),
    max = max(hourly_counts, na.rm = TRUE)
  )
```

Those are the parts of the box: the line is the median, the box edges are the
quartiles, and the whiskers reach out to 1.5x the interquartile range (IQR) - which is quartile 1  - 3. Things beyond the whiskers are classed as outliers. 

::: {.callout-tip title="Going deeper: the box is more adjustable than it looks" collapse="true"}

Boxplots have more calculation options than you might realise. You can change
how the box width is computed, and how outliers are decided.

If you want to read more, look up `?geom_boxplot()` and `boxplot.stats`.

:::

So, `geom_boxplot()` does some computation, to return these summaries, and then draw them. So why is the 8am box so wide? The median is 4910 but the lower quartile is 233, which means at least a quarter of the mornings had almost nobody there.

:::{.callout-note title="Your Turn"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs", "flagstaff-data"]
#| eval: false
#| echo: true
```

:::
Why do you think there is so much variation across 8am? How can it be so busy, and so quiet, at 8am?

How would you go about exploring this? 

Would you plot something? Would you do some data summarising? 

What would you look at?

How would you plot it? What do you expect?

Let's take a couple of minutes to explore this, draw some sketches, imagine the types of data summary we might see as tables, as plots!

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```


:::

Let's explore this by looking at each weekday:

```{r}
#| label: eight-am-split

flagstaff_8am |> 
  group_by(week_day) |>
  summarise(median_count = median(hourly_counts, na.rm = TRUE))
```

Nobody catches a train to work on a weekend, and the boxplot summarised across every day at once - so it generalised across both.

We come back to this properly in @sec-in-the-way. For now, notice that the
computation was correct, even if the the answer was a little misleading.

::: {.callout-note title="Your Turn"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs", "flagstaff-data", "five-number"]
#| eval: false
#| echo: true
```

:::
Here is the plot again, so you have something to change:

```{r}
#| eval: false
ggplot(flagstaff,
       aes(x = hour,
           y = hourly_counts,
           group = hour)) +
  geom_boxplot()
```

1. Swap `group = hour` for `x = factor(hour)`. Do you get the same plot? Which
   version do you prefer, and what changed on the x axis?

2. Which other hours have suspiciously tall boxes? Can you explain them?

3. We worked out that 8am is really two mornings. Draw it as two boxes - complete the code below:

```{r}
#| eval: false
flagstaff_8am_weekend <- flagstaff_8am |> 
  mutate(
    day_type = if_else(
      condition = week_day %in% c(..., ...),
      true = ...,
      false = ...
    )
  )

flagstaff_8am_weekend

  ggplot(...,
         aes(x = ...,
             y = ...)) +
  geom_boxplot()
```

Is that a fairer picture of 8am than the one box was?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you have actually had a go.

**1.** Same boxes. The difference is the x axis: `group = hour` keeps it a
number, so the boxes sit at 0, 1, 2 and the spacing is real. `factor(hour)`
makes it a category, so they are evenly spaced and every hour gets a label. For
hours I prefer `group`, because the axis still means something.

**2.** The four widest boxes are hours 8, 17, 9 and 7, which is the morning and
evening commute. Those are the hours where weekday and weekend differ most, so
those are the hours where one box is hiding two populations.

**3.** Much fairer. Two boxes, each describing a kind of morning that actually
happens.

:::

:::

Boxplots are a really useful tool for exploring variation. Sometimes the wider variation can have distinct and interesting summaries.

### Takehomes

A boxplot turns however many numbers you have into five, and draws those.

That is a good trade when the data has one lump in it. When it has two, the five
numbers describe a middle that nothing is near.

## Geoms that count for you

> **The idea:** some geoms draw the number you give them. Others work a number
> out for you. If you do not know which kind you have, you will get a plot of
> something you did not ask for.

The boxplot kept all the variation. Sometimes you just want one number per hour, and you want to say which number.

The number of people is in `hourly_counts`, but there's one row per day per
hour, and I want one number per hour. So I summarise first.

Data summarising is an important skill, so let's take a minute to sketch out what the data summary would look like.

:::{.callout-note title="Your Turn: group_by(), summarise()"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs", "flagstaff-data", "flagstaff-hourly"]
#| eval: false
#| echo: true
```

:::
Sketch out the data as a table for the summary we want - which is:

- One row per hour
- Each hour has the mean number of counts in that hour

Sketch it out!

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 2)
```


It should look something like the following:

<details>
<summary>Answer</summary>

```{r}
tibble(
  hour = c("0", "1", "...", "23"),
  mean_count = c("245", "125", "...", "180")
)
```

</details>


```{r}
#| label: flagstaff-hourly
flagstaff_hourly <- flagstaff |>
  group_by(hour) |>
  summarise(mean_count = mean(hourly_counts, na.rm = TRUE))

flagstaff_hourly
```

Twenty four rows: one row per hour, and a summary.

That is exactly what `geom_col()` wants. You give it an x and a y, and it draws
a bar that tall.

```{r}
#| label: fig-flagstaff-col
#| fig-alt: |
#|   A bar chart of mean pedestrian count by hour at Flagstaff Station. Two tall
#|   peaks, one at 8am reaching about 3500 and one around 5pm, with very low
#|   bars overnight.
#| fig-cap: "Flagstaff Station. A train station, drawn as a train station."
ggplot(flagstaff_hourly,
       aes(x = hour,
           y = mean_count)) +
  geom_col()
```

A spike at 8am, another at 5pm, almost nobody overnight.

It is only describing a single number - not the variation.

This doesn't make `geom_col()` wrong! You've displayed a summary.

::: {.callout-tip title="So what does `geom_bar()` do?" collapse="true"}

There are two bar geoms, and the other one is `geom_bar()`.

The difference is that `geom_col()` needs you to give it a y, and `geom_bar()`
works one out for itself. What it works out is a **count of rows**.

Watch what happens if I use it on the raw data.

```{r}
#| label: fig-flat-wall
#| fig-alt: |
#|   A bar chart of hour on the x axis. All twenty four bars are the same
#|   height, about 1570, making a flat wall.
#| fig-cap: "Every hour is exactly as busy as every other hour. Suspicious."
ggplot(pedestrian,
       aes(x = hour)) +
  geom_bar()
```

Every hour of the day is equally busy, and Melbourne at 3am is as busy as
Melbourne at 5pm. Strange?

`geom_bar()` doesn't take a y value, instead it counts how many **rows** landed in
each value of `hour`, which is this:

```{r}
#| label: count-hour
pedestrian |>
  count(hour)
```

About 1570 every hour. The plot is drawing my data collection, not my pedestrians - I haven't given it the right info.

So `geom_bar()` is just `count()` and then `geom_col()`, glued together. These
two are the same plot:

```r
ggplot(pedestrian, aes(x = hour)) +
  geom_bar()

pedestrian |>
  count(hour) |>
  ggplot(aes(x = hour, y = n)) +
  geom_col()
```

That can be a useful thing to do, when rows are the thing you want counted. This would be a useful thing to go to if each row were one person walking past a sensor. Instead, each row here is an hour of recording, so counting rows tells me about the sensor rather than the street.


So the two of them, side by side:

> `geom_col()` uses the number you already have.
>
> `geom_bar()` counts your rows for you, the same way `count()` does.

Rather than memorising that, I'd ask a question about your data instead:

> Have I already worked out the number I want on the y axis?

If yes, `geom_col()`. If no, and rows are the thing you want counted,
`geom_bar()`.

:::


::: {.callout-note title="Your Turn"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs", "flagstaff-data", "flagstaff-hourly"]
#| eval: false
#| echo: true
```

:::
1. Before you run anything, sketch what you think Birrarung Marr looks like
   across the day. It's parkland by the river. Where do you think the peak is?

2. Now build it. Fill in the blanks:

```{r}
#| eval: false
birrarung_hourly <- pedestrian |>
  filter(sensor_name == "Birrarung Marr") |>
  group_by(___) |>
  summarise(mean_count = mean(___, na.rm = TRUE))

ggplot(birrarung_hourly,
       aes(x = ___,
           y = ___)) +
  geom_col()
```

   How close was your sketch?

3. Draw the park as a boxplot too, the same way we did for Flagstaff. Does the
   park have an 8am problem like the station did, or is the mean a fair summary
   here?

4. Now swap `geom_col()` for `geom_bar()` on `birrarung_hourly`. **You should get an error.** Read it, and see whether you can explain it from what
   `geom_bar()` is trying to do.

5. What happens if you drop the `na.rm = TRUE` from the `mean()`?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you have actually had a go.

**2.** The park peaks at 5pm, not 8am. People walk by the river after work,
not on the way to it.

```{r}
#| eval: false
birrarung_hourly <- pedestrian |>
  filter(sensor_name == "Birrarung Marr") |>
  group_by(hour) |>
  summarise(mean_count = mean(hourly_counts, na.rm = TRUE))

ggplot(birrarung_hourly,
       aes(x = hour,
           y = mean_count)) +
  geom_col()
```

**3.** No 8am problem. A park does not care what day it is, so the mean is a
fairer summary here than it was at Flagstaff.

**4.** You get `Problem while computing stat.` `geom_bar()` wants to count rows
and work out `y` itself, and you have already given it a `y`. It cannot do both.

**5.** Every hour comes back `NA`, all 24 of them. One missing reading anywhere
in an hour is enough to poison the whole mean.

:::

:::

### Takehomes

`geom_col()` draws the number you give it. `geom_bar()` counts your rows and
draws that.

So the question is not "which function is right", it is **have I already worked
out the number I want on the y axis?**

## Geoms that bin for you

> **The idea:** a histogram chops a continuous variable into bins and counts
> what lands in each. The bin width is your choice, and it decides what shape
> you see.

:::{.callout-note title="Your Turn: Do you know what binning is?"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs"]
#| eval: false
#| echo: true
```

:::
Let's chat about binning! Do you know what a histogram does?

Let's explore what `hist()` does:

- **demo**

:::

::: {.callout-note title="Your Turn: draw it first"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs"]
#| eval: false
#| echo: true
```

:::
We are about to look at the spread of `hourly_counts` across all four sensors.
Every hour of 2016, at four places, as one distribution.

Draw the shape you expect:

- Number of people in an hour
- How often that happens

- Is it symmetric, or lopsided?
- is the tall part on the left or the right?

:::

Let's look at the spread of `hourly_counts` across all four sensors, using `geom_histogram()`

```{r}
#| label: fig-hist-default
#| fig-alt: |
#|   A histogram of hourly pedestrian counts. One very tall bar at the left
#|   holding most of the data, and a long thin tail stretching right to about
#|   11000.
#| fig-cap: "The default. Note what ggplot2 says about it."
ggplot(pedestrian,
       aes(x = hourly_counts)) +
  geom_histogram()
```

Two messages came out with that plot, let's read them:

```
`stat_bin()` using `bins = 30`. Pick better value `binwidth`.
```

ggplot2 picked 30 bins because it had to pick something, and it is telling you
that 30 was not a considered choice. It's a placeholder, and it would like you to
replace it.

The second is our old friend from @sec-anatomy:

```
Removed 2548 rows containing non-finite outside the scale range
```

There are 2548 hours where the sensor recorded nothing. We come back to those in
@sec-in-the-way.

So is 30 bins any good here? With a range of 0 to 11273, thirty bins makes each
one about 389 people wide, and the first bin alone holds 40% of the data.

So most of the picture ends up as one bar.

Let's try some other widths.

:::{.panel-tabset}

## binwidth = 1000

```{r}
#| label: fig-bw-1000
#| warning: false
#| fig-alt: A histogram with wide bins, showing about eleven bars.
ggplot(pedestrian, aes(x = hourly_counts)) +
  geom_histogram(binwidth = 1000)
```

## binwidth = 100

```{r}
#| label: fig-bw-100
#| warning: false
#| fig-alt: |
#|   A histogram with narrower bins, showing a smooth decay from a peak near
#|   zero.
ggplot(pedestrian, aes(x = hourly_counts)) +
  geom_histogram(binwidth = 100)
```

## binwidth = 10

```{r}
#| label: fig-bw-10
#| warning: false
#| fig-alt: |
#|   A histogram with very narrow bins, spiky and noisy, with a tall cluster
#|   near zero.
ggplot(pedestrian, aes(x = hourly_counts)) +
  geom_histogram(binwidth = 10)
```

:::

At 1000 you get the shape and nothing else. At 10 you get every wobble and no
shape. At 100 I can see both, so that's the one I would keep.

How do you choose? It depends! You should explore

The question I find useful is **what is a meaningful unit here?** A binwidth of
100 people an hour is a number I can describe to somebody. A binwidth of 388.7
is not a number anybody chose.

### Back to that 8am box

Now we've got a tool that can answer the question the boxplot raised.

The 8am box ran from 233 to 5128, and I said that meant 8am was not one thing.
A histogram of just those mornings should show us what it actually is.

```{r}
#| label: fig-hist-8am
#| fig-alt: |
#|   A histogram of pedestrian counts at Flagstaff Station at 8am. Two clearly
#|   separated humps, one near zero and a much larger one around 5000, with an
#|   almost empty gap between them.
#| fig-cap: "8am at Flagstaff Station. Two humps, not one."

ggplot(flagstaff_8am,
       aes(x = hourly_counts)) +
  geom_histogram(binwidth = 50)
```

There are the two types of mornings.

One hump down near zero, one hump up around 5000, and almost nothing in
between.

The boxplot couldn't show you this. A histogram (with good bin size!) has as many humps as the data has, and here the data has two.

### Showing the actual readings with `geom_rug()`

A histogram is still a summary, though. Those bars are counts of things, not the
things themselves.

`geom_rug()` draws a small tick for every single observation, along the axis. So you
can put the summary and the raw data on the same plot, which I think is lovely.

```{r}
#| label: fig-rug
#| fig-alt: |
#|   The same 8am histogram with a rug of tick marks along the bottom axis. The
#|   ticks are dense under each of the two humps and almost absent in the gap
#|   between them.
#| fig-cap: "Every one of the 397 mornings, drawn underneath the bins."
ggplot(flagstaff_8am,
       aes(x = hourly_counts)) +
  geom_histogram(binwidth = 50) +
  geom_rug(alpha = 0.3)
```

Look along the bottom. Each tick is one morning in 2016, and you can see them
bunch under the two humps and thin out to nearly nothing across the middle.

I like `geom_rug()` a lot for this - because it costs you one line and it puts
the evidence next to the summary.

::: {.callout-tip title="Comparing two distributions on one plot" collapse="true"}

The other thing you often want is two distributions side by side. Let's take the
station and the park.

```{r}
#| label: two-sensors
two_sensors <- pedestrian |>
  filter(sensor_name %in% c("Flagstaff Station", "Birrarung Marr"))
```

The obvious move is to map the sensor onto `fill`, and make them transparent so
we can see through one to the other.

```{r}
#| label: fig-overlay-stacked
#| warning: false
#| message: false
#| fig-alt: |
#|   A histogram of two sensors with semi transparent fill. The bars sit on top
#|   of each other rather than overlapping, reaching about 7400 at the tallest.
#| fig-cap: "Transparent, and still wrong."
ggplot(two_sensors,
       aes(x = hourly_counts,
           fill = sensor_name)) +
  geom_histogram(binwidth = 100, alpha = 0.5)
```

That's not two overlapping distributions - the are stacked on top of each other.

`alpha` made them see-through, but see-through was never the problem. The
problem is **position**. By default `geom_histogram()` stacks, so the second
sensor's bars start where the first sensor's bars finish.

What we want is for both to start at zero, and that is `position = "identity"`.

```{r}
#| label: fig-overlay-identity
#| warning: false
#| message: false
#| fig-alt: |
#|   The same two histograms, now both starting from zero and overlapping, so
#|   the park's distribution sits inside the station's.
#| fig-cap: "Both starting from zero. Now they overlap."
ggplot(two_sensors,
       aes(x = hourly_counts,
           fill = sensor_name)) +
  geom_histogram(binwidth = 100,
                 alpha = 0.5,
                 position = "identity")
```

Now you can read it. Birrarung Marr sits low and tight, because a park is
quietly busy most of the time. Flagstaff has a long tail out to the right,
because a train station is empty at 3am and enormous at 8am.

If the overlap still bothers you, `geom_density()` does a related computation and
draws a smooth curve instead of bars. I like to pair `geom_density()` with `geom_rug()`:

```{r}
#| label: fig-density
#| warning: false
#| fig-alt: |
#|   Two smooth density curves for the same two sensors, overlapping, with the
#|   park's curve narrow and the station's curve long tailed.
#| fig-cap: "The same comparison, as curves."
ggplot(two_sensors,
       aes(x = hourly_counts,
           colour = sensor_name)) +
  geom_density() +
  geom_rug(alpha = 0.2)
```

Be careful with density curves, though. That y axis is no longer a count of
anything. It is a density, chosen so the area under each curve is one, which is
what lets you compare two groups of different sizes.

:::

::: {.callout-note title="Your Turn"}


::: {.callout-tip title="Setup code" collapse="true"}

```{r}
#| ref.label: ["pkgs", "flagstaff-data", "five-number"]
#| eval: false
#| echo: true
```

:::
1. Draw the 8am histogram for **Birrarung Marr** before you build it. One hump
   or two? A park doesn't much care what day it is, so what should that mean?

2. Now build it, with a rug:

```{r}
#| eval: false
birrarung_8am <- pedestrian |>
  filter(sensor_name == "Birrarung Marr",
         hour == ___)

ggplot(birrarung_8am,
       aes(x = hourly_counts)) +
  geom_histogram(binwidth = ___) +
  geom_rug(alpha = ___)
```

   Were you right? Try a couple of other binwidths while you are there.

3. Go back to `flagstaff_8am` and try `binwidth = 2000`. What happens to the two
   humps, and what would you have concluded if you had only ever seen that
   version?

4. `geom_histogram()` bins for you. `geom_freqpoly()` does the same computation
   and draws a line instead. Swap one for the other on `flagstaff_8am`.

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you have actually had a go.

**1 and 2.** One hump. Birrarung Marr at 8am runs from 95 to 7218 with no gap in
the middle, because a park does not care what day it is. No commute means no
second population.

**3.** At `binwidth = 2000` you get four bars, and the two humps merge into one.
You would conclude 8am at Flagstaff is a single busy-ish morning, which is
exactly backwards.

**4.** Same computation, different mark. `geom_freqpoly()` is easier to overlay
two of, and harder to read a single count off.

:::

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

:::

### Takehomes

The default binwidth is a placeholder, and ggplot2 says so out loud. Pick a
number you could describe to somebody.

A histogram is still a summary, so put a `geom_rug()` under it when there are
few enough points to see them.

## To summarise

Three things to take out of this chapter.

1. **Some geoms compute, and the computation is a choice you can see.** If you
   didn't give it a y, it made one. If it drew a box, it worked out five numbers
   and threw the rest away.
2. **The default binwidth is a placeholder,** and ggplot2 says so out loud. Pick
   a unit you could describe to somebody, and put a `geom_rug()` under it so you
   can see what you are summarising.
3. **A summary is a choice, not a fact.** The boxplot's five numbers, the mean
   in a bar, the height of a bin. Each one throws something away, and it is
   worth knowing what.

And the habit underneath all three.

When a plot surprises you, ask what got computed between the data and the marks.
Nearly every surprise in this chapter came from a computation happening exactly
as documented, on data I hadn't noticed was different to what I thought.

Next up is what you are comparing, which is where we stop asking how a plot is
built and start asking what it is for.
