# Where should the eye go? {#sec-eye}

@sec-in-the-way was about taking things out. This chapter is about what's left,
and which part of it a reader meets first.

A plot where everything has equal weight has no first, second or third, so the
reader has to do the ranking themselves.

You're spending their attention whether you decide to or not.

::: {.overview}

## Overview {.unnumbered}

**Duration** 50 minutes

**Questions**

<!-- These map onto the sections below, in order. If you move a section, move its question. -->

- What does a reader look at first?
- How do I get emphasis for free?
- How do I choose colours that every reader can tell apart?
- How do I make one group stand out against the others?

**What you need this session**

- A session of RStudio open
- Something to draw on, and something to draw with
- The following packages installed

```{r}
#| label: pkgs
#| message: false
#| output: false
library(tidyverse)
library(naniar)
library(colorspace)
library(gghighlight)
library(ggrepel)
```

```{r}
#| label: pkgs-extra
#| echo: false
#| output: false
library(countdown)
```

```{r}
#| label: ch7-data
#| message: false
pedestrian_hourly <- pedestrian |>
        mutate(day_type = if_else(
                condition = str_starts(week_day, "S"),
                true = "weekend",
                false = "weekday"
                )
        ) |>
        group_by(sensor_name, day_type, hour) |>
        summarise(
                mean_count = mean(hourly_counts, na.rm = TRUE),
                .groups = "drop"
        )

sensor_totals <- pedestrian |>
        group_by(sensor_name) |>
        summarise(total = sum(hourly_counts, na.rm = TRUE))
```

:::

## Readers don't scan evenly

> **The idea:** the channels you map data onto are not equally easy to read.
> Some are precise, some are rough, and the order is known.

Back in @sec-anatomy I said position beats colour beats shape beats size, and
asked you to take it on trust.

Here's where that comes from. Cleveland and McGill ran experiments in 1984 on
how accurately people read each channel, and got a ranking:

1. position on a common scale
2. position on non-aligned scales
3. length
4. angle
5. area
6. colour saturation

Which is why:

- a bar chart is easier to read than a pie chart, because length beats angle
- mapping something important to `size` is usually a mistake
- facets cost you accuracy, because each panel is its own scale

Put your most important variable on a position, every time.

::: {.callout-note title="Your Turn"}

1. Look back at the `size` and `shape` plots in @sec-anatomy. Given the ranking,
   would you make the same choices now?

2. Which channel is doing the most work in the last plot you made at work?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 3)
```

:::

### What to hold onto

- Position is the most accurate channel, colour and area are among the worst
- The most important variable goes on a position
- Everything after that is a compromise you're choosing on purpose

## Order is free emphasis

> **The idea:** the order of your categories is a decision. If you don't make
> it, the alphabet makes it for you.

Our four sensors have been alphabetical since @sec-comparing.

```{r}
#| label: sensor-totals-print
sensor_totals |>
        arrange(desc(total))
```

Alphabetical order puts the third busiest first and the busiest second. The
letter B is doing work that traffic should be doing.

:::{.panel-tabset}

## alphabetical

```{r}
#| label: fig-alpha-order
#| fig-alt: |
#|   A bar chart of total foot traffic per sensor, in alphabetical order, so the
#|   bar lengths jump up and down with no pattern.
ggplot(sensor_totals,
       aes(x = total,
           y = sensor_name)) +
        geom_col()
```

## sorted

```{r}
#| label: fig-sorted
#| fig-alt: |
#|   The same bar chart sorted by total, so the bars form a clean descending
#|   staircase.
ggplot(sensor_totals,
       aes(x = total,
           y = fct_reorder(sensor_name, total))) +
        geom_col()
```

:::

One function, and the plot went from a lookup table to a ranking.

The same works on facets, so the panels read in order:

```{r}
#| label: fig-facet-sorted
#| fig-alt: |
#|   Four panels of hourly foot traffic, ordered from busiest sensor to
#|   quietest rather than alphabetically.
pedestrian_hourly |>
        mutate(sensor_name = fct_reorder(sensor_name, -mean_count, .fun = max)) |>
        ggplot(aes(x = hour,
                   y = mean_count,
                   colour = day_type)) +
        geom_line() +
        facet_wrap(vars(sensor_name))
```

Sorting is usually a good idea. Not always:

- **sort** when the reader is comparing
- **don't sort** when the reader is looking something up

A phone book sorted by height would be a nightmare.

::: {.callout-note title="Your Turn"}

1. Add `fct_rev()` around the `fct_reorder()`. Which direction reads better for
   a horizontal bar chart?

2. Sort the facets by **weekend** traffic instead of overall traffic:

```{r}
#| eval: false
pedestrian_hourly |>
        filter(day_type == "___") |>
        mutate(sensor_name = fct_reorder(sensor_name, ___))
```

   Does the story change?

3. `fct_infreq()` sorts by how often a value appears. When would you want that
   instead of `fct_reorder()`?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** `fct_rev()` puts the biggest at the top. For a horizontal bar chart I
prefer that, because English readers start top left and the most important bar
is the first thing they meet.

**2.** `filter(day_type == "weekend")` then `fct_reorder(sensor_name,
mean_count)`. The order changes: Birrarung Marr climbs, because a park is a
weekend place.

**3.** `fct_infreq()` counts rows, so use it when the thing you care about *is*
how often a category turns up. Bar charts of raw counts, mostly.

:::

:::

### What to hold onto

- Alphabetical is not an order, it's the absence of one
- `fct_reorder()` on a variable your reader cares about is the cheapest
  emphasis available
- Sort for comparison, don't sort for lookup

## Colour that carries meaning

> **The idea:** colour is the channel most likely to fail silently. It can look
> fine to you and be unreadable to one reader in twelve.

Here's our four sensors in ggplot2's default colours.

```{r}
#| label: fig-default-cols
#| fig-alt: |
#|   Four hourly foot traffic lines, one per sensor, in ggplot2's default red,
#|   green, blue and purple.
p_sensors <- ggplot(pedestrian_hourly,
                    aes(x = hour,
                        y = mean_count,
                        colour = sensor_name)) +
        geom_line()

p_sensors
```

Those four are `#F8766D`, `#7CAE00`, `#00BFC4`, `#C77CFF`. Here's what they
become under deuteranopia, the commonest kind of colour vision deficiency:

```{r}
#| label: cvd-default
deutan(scales::hue_pal()(4))
```

Two olives and two blues. And on the plot:

```{r}
#| label: fig-cvd-sim
#| fig-alt: |
#|   The same four lines with the colours simulated for deuteranopia, where two
#|   pairs are now nearly identical.
p_sensors +
        scale_colour_manual(values = deutan(scales::hue_pal()(4)))
```

Half the information is gone. This is roughly 8% of men of northern European
descent, and about 0.5% of women.

**And the palette you were about to reach for has the same problem.** Dark2 is
the one everyone uses:

```{r}
#| label: cvd-dark2
deutan(RColorBrewer::brewer.pal(4, "Dark2"))
```

The first and fourth are near identical greys.

Viridis survives, because it varies **lightness** as well as hue:

```{r}
#| label: cvd-viridis
deutan(viridisLite::viridis(4))
```

Dark to light, still in order.

```{r}
#| label: fig-viridis
#| fig-alt: |
#|   The same four lines in the viridis palette, which runs dark purple through
#|   green to yellow.
p_sensors + scale_colour_viridis_d()
```

### Match the palette to the variable

Most bad colour is a type mismatch:

- **qualitative** for unordered categories, like our four sensors
- **sequential** for magnitude, like a count
- **diverging** for distance from a midpoint, like above and below average

::: {.callout-tip title="Read more"}

Two tools for checking a palette before you commit to it:
[Quickly Assessing Colour
Palettes](https://www.njtierney.com/posts/2020-10-15-assess-colour/).

`prismatic::check_color_blindness()` simulates a set of hex codes under each
kind of deficiency, and `colorspace::demoplot()` paints a palette into a fake
plot so you can judge it without going through your own data.

:::

::: {.callout-tip title="Going deeper: hour is cyclic" collapse="true"}

`hour` is none of the three types. Midnight is next to 23:00, so a sequential
palette puts the two closest hours at opposite ends of the scale.

There's no good default for this. The three way taxonomy is useful and it isn't
the whole story.

:::

::: {.callout-note title="Your Turn"}

1. Run `deutan()` on a palette you use at work. Does it survive?

2. Simulate a different brewer palette:

```{r}
#| eval: false
p_sensors +
        scale_colour_manual(
                values = deutan(RColorBrewer::brewer.pal(4, "___"))
        )
```

   Try `"Set1"` and `"Set2"`. Better or worse than Dark2?

3. Colour the plot by `mean_count` instead of `sensor_name`. Which kind of
   palette does ggplot2 give you, and is it the right kind?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**2.** Set1 holds up better than Dark2 because its colours differ in lightness
as well as hue. Set2 is pastel and collapses badly.

**3.** A continuous, sequential palette, because `mean_count` is a number. That
is the right *kind*, and it's still the wrong choice here: you have four
distinct places, not a magnitude.

:::

:::

### What to hold onto

- The default palette fails under colour vision deficiency, and so does Dark2
- Palettes that vary lightness survive, which is why viridis is the safe default
- Simulate it with `deutan()` rather than guessing

## Emphasis by contrast

> **The idea:** four colours asks the reader to rank four things. Usually you
> want them to look at one.

Grey is not an absence of colour. It's a demotion, and you do it on purpose.

```{r}
#| label: fig-demote
#| fig-alt: |
#|   Four hourly lines, three in pale grey and one, Flagstaff Station, in dark
#|   red.
ggplot(pedestrian_hourly,
       aes(x = hour,
           y = mean_count,
           group = sensor_name)) +
        geom_line(colour = "grey80") +
        geom_line(data = filter(pedestrian_hourly,
                                sensor_name == "Flagstaff Station"),
                  colour = "firebrick")
```

Two `geom_line()` calls, one of them filtered. `gghighlight()` from
@sec-in-the-way does the same job in one line, and it's worth having seen the
manual version so it isn't magic.

Then get rid of the legend by putting the label on the line.

```{r}
#| label: fig-direct-label
#| warning: false
#| fig-alt: |
#|   The four sensor lines with each name printed at the right hand end of its
#|   own line, and no legend.
sensor_labels <- pedestrian_hourly |>
        group_by(sensor_name) |>
        slice_max(hour, n = 1)

p_sensors +
        geom_text_repel(data = sensor_labels,
                        aes(label = sensor_name),
                        hjust = 0,
                        direction = "y",
                        nudge_x = 1) +
        theme(legend.position = "none") +
        coord_cartesian(xlim = c(0, 32))
```

A legend makes the eye bounce between the key and the plot, once per series. A
label sits where the thing is.

::: {.callout-note title="Your Turn"}

1. Run the greyed plot through `deutan()`. Does highlighting survive colour
   vision deficiency better than four categorical hues did?

2. Highlight a different sensor:

```{r}
#| eval: false
ggplot(pedestrian_hourly,
       aes(x = hour, y = mean_count, colour = sensor_name)) +
        geom_line() +
        gghighlight(sensor_name == "___")
```

3. If the question were "where is busy on a weekend", which sensor would you
   highlight? Is that the same answer you gave in @sec-comparing?

```{r}
#| echo: false
#| eval: !expr knitr::is_html_output()
countdown(minutes = 5)
```

::: {.callout-tip title="Answer" collapse="true"}

Only open this if you've actually had a go.

**1.** Yes. Grey against one colour is a lightness difference, and lightness is
the thing that survives. This is the strongest argument for highlighting over
four hues.

**3.** Birrarung Marr, because it's the only sensor where the weekend beats the
weekday. Same answer as @sec-comparing, which is a good sign: the emphasis
should follow the finding, not replace it.

:::

:::

### What to hold onto

- Grey is a decision, not a default
- Highlight one series rather than colouring four
- A direct label beats a legend, because the reader stops travelling

## To summarise

Three things to take out of this chapter.

1. **Position is the most accurate channel.** Spend it on what matters, and
   treat colour and area as rough.
2. **Order is free emphasis.** `fct_reorder()` costs one function and turns a
   lookup table into a ranking.
3. **Colour fails silently.** Simulate it, and prefer palettes that vary
   lightness.

And the habit: before you style anything, decide what you want the reader to see
first. Then make that the loudest thing on the page and demote everything else.

Next up is making it land, which is the last mile: labels, themes, and getting a
file out that looks the way it did on your screen.
